---
layout: page
title: IntermediateR for reproducible scientific analysis
subtitle: data.table
minutes: 20
---


> ## Learning objectives {.objectives}
>
> * To know how to perform common data.frame tasks using data.table
> * To know how to examine the structure of objects in R
> * To be able to set keys for data.table
>

First, we will be learning about the `data.table` package. Data tables
have a number of advantages over data frames:

 * It provides a huge speed increase over data frames. Tasks that would
   normally take hours with a data frame take seconds with data tables.
 * It is much more concise to write and allows you to avoid complicated
   which statements and repeated subsetting.
 * Its key system allows you to merge/query multiple tables without 
   worrying about matching rownames.

### Reading in data

Let's load in the package and read in some data:

```{r}
library("data.table")
gap <- fread("data/gapminder-FiveYearData.csv")
gap
```

We can see the data has loaded correctly using the `fread` function from the 
data.table package. Note that unlike data frames, R will print out the first and 
last 5 rows ofthe table.

`fread` works similary to `read.table`: it tries to make
sense of the data and read it in appropriately. It is much faster than 
`read.table` for large tables, but is slightly less sophisticated. You may find
yourself needing to use `read.table` or one of its derivative functions to load
in data correctly, then casting to a data table:

```{r}
# Note that as.data.table will throw out rownames unless you set 
#'keep.rownames = TRUE'
gap_df <- read.csv("data/gapminder-FiveYearData.csv")
gap_dt2 <- as.data.table(gap_df)
```

We can use the structure function (`str`) to examine what exactly a data.table 
is:

```{r}
str(gap)
```

A data.table is simply a data.frame with an additional class attached to it, 
along with a special attribute, ".internal.selfref", which is an external pointer
that data.table uses to work with the data in memory in a lower level language.

Data tables are backwards compatible with most functions that require data 
frames, however you may occasionally find that you need to cast them using
`as.data.frame` on functions which check only the first class of the object.

To prove this to ourself, we can check the objets for equality:

```{r}
all.equal(gap, gap_df)
all.equal(gap, gap_df, check.attributes = FALSE)
```

This shows us that the only the attributes of the object are different, but the
underlying data is the same.

### Basic operations

Nearly all operations on data tables are performed inside the `[` function, and
are performed in place in memory. Let's take a look at some operations and their
data frame equivalents:

To select or filter rows, we use the first argument of `[` just like with data
frames:

```{r}
gap[continent == "Oceania"]
# data frame equivalent
gap_df[gap_df$continent == "Oceania",]
```

The data table knows when we type `continent` to look for that as a column in
the table: removing the redundant text we'd need to type for data.frames. This 
becomes even more convenient when filtering on multiple columns:

```{r}
gap[continent == "Oceania" & country == "Australia" & year %in% c(1952, 2007)]
# data frame equivalent
gap_df[gap_df$continent == "Oceania" & gap_df$country == "Australia" & gap_df$year %in% c(1952, 2007),]
```

Note that the data frame call will still work on the data table:

```{r}
gap[gap$continent == "Oceania" & gap$country == "Australia" & gap$year %in% c(1952, 2007),]
```

To select columns we use the second argument to `[`, just like data.frames:

```{r}
gap[,continent]
# data frame equivalent
gap_df[,"continent"]
```

To select multiple columns, we need to pass in the column names as a list:

```{r}
gap[, list(continent, country, pop)]
# data frame equivalent
gap_df[, c("continent", "country", "pop")]
```

We can also rename columns in the output using the list:

```{r}
gap[, list(a=continent, b=country, c=pop)] # This does not alter the data table
```

We can create temporary columns, those that only exist in the output data 
structure using list arguments:

```{r}
# total_gdp only exists in the output
gap[,list(continent, country, year, total_gdp=pop*gdpPercap)]
# Lets see what gap contains again:
gap
# The equivalent data frame call:
cbind(gap_df[,c("continent", "country", "year")], total_gdp = gap_df$pop * gap_df$gdpPercap)
```

To create a new column in the data table, we have to use the special operator 
`:=`:

```{r}
# Add a new column to the gapminder data with total gdp
gap[, total_gdp := gdpPercap * pop]
# data frame equivalent
gap_df <- cbind(gap_df, total_gdp = gap_df$gdpPercap * gap_df$pop)
```

### Data table specific operations

Data tables have a number of special variables that are useful in calculations:

```{r}
# get all the columns
gap[,.SD]
# get the number of rows
gap[,.N]
# generate indices for the rows
gap[,.I]
```

These become useful in conjuction with data table's extra arguments to `[`:

```{r}
# How many countries in each continent?
gap[year == 2007, list(countries=.N), by=continent]
```

The `by` argument lets you calculate things within groups:

```{r}
# Mean life expectancy per continent per year:
gap[, list(avgLifeExp=mean(lifeExp)), by=list(continent, year)]
```

The `with` argument lets you pass in column names as a character vector:

```{r}
gap[,c("continent", "country", "year"), with=FALSE]
```

### Keys 

One of the advantages of data table is the ability to set each tables "keys":
the columns which will act as unique identifiers for each row, for example:

```{r}
setkey(gap, continent, country, year)
```

We can see the change using the `tables` function, which shows all data tables 
in the R session:

```{r}
tables()
```

This is really useful when you have multiple tables: allowing you to efficiently
merge tables together, confidently and concisely:

```{r}
landSize <- data.table(
  country=c("Australia", "New Zealand"),
  size=c(7692024, 268021)
)
setkey(landSize, country)
setkey(gap, country)
# Join landSize to gap, keeping only 'keys' (rows) in gap that also exist in
# landSize
gap[landSize]
# And vice-versa:
landSize[gap]
```

To learn more about data table, you can check out the package Vignette on CRAN:
<http://cran.r-project.org/web/packages/data.table/vignettes/datatable-intro.pdf>




