---
layout: page
title: Intermediate R for reproducible scientific analysis
subtitle: reshape2
minutes: 20
---


> ## Learning objectives {.objectives}
>
> * To know how split a column by a group to create multiple columns
> * To know how to combine multiple columns into one with different groupings.
>

So far we've been working with the gapminder dataset:

```{r}
library("data.table")
gap <- fread("data/gapminder-FiveYearData.csv")
gap
```

The format that this dataset has been provided in is called a "long" format:
each variable has its own column, and there are several columns with identifying
information: where those observations came from/groupings in the data.

This is a format that is convenient for rapid analysis using R: it is easy to
manipulate, filter and query using `data.table`, and is the format `ggplot2`
expects for plotting.

However, data often does not come in this format. We might sometimes find we are
given data in what is known as a "wide" format: in which variables may be split
over multiple columns, one for each group. Let's read in an alternative version
of the gapminder dataset:

```{r}
# This is an example where `fread` doesn't work: it loses the column names!
gapWide <- as.data.table(read.csv("data/gapminder-wide-format.csv", header=TRUE))
gapWide
```

This contains exactly the same data, but the three variables, "gdpPercap", "pop",
and "lifeExp" have one column for each year of collection. This format is useful
if you wish to calculate group-wise properties, e.g. the correlation structure
between groups. It is also a more natural way for data collectors to structure
data in another application, such as Excel.

Being able to transform between "wide" and "long" formats is an incredible useful
skill that will save you a lot of time in the future. We can do this using the
`reshape2` package.

### Combining columns

To collapse multiple columns into one we use the `melt` function:

```{r}
library(reshape2)
gapLong <- melt(
  data=gapWide,
  id.vars=c("continent", "country") # All other columns will be collapsed into one
)
```

We get a warning because the 'pop' columns are of type 'integer' (i.e. whole 
numbers) while the 'gdpPercap' and 'lifeExp' columns are type 'double' (i.e. 
decimal values).

This melt has collapsed the table too far. We need to separate out the different
variable types into their own columns. First, lets split the variable column into
variable type and year:

```{r}
gapLong[, c("var", "year") := colsplit(variable, "\\.", c("var", "year"))]
```

Let's break this down.

First, `colsplit(variable, "\\.", c("var", "year"))` creates two columns called 
"var" and "year" from the "variable" column, splitting each value on the ".".
We need to specific the pattern as `\\.` because `.` by itself is a wild card
character: it would match every character in the string.

Next, we create two new columns in the `gapLong` data table using the `:=` 
operator. To create multiple columns, we specify their names in a vector to the
left of the `:=` operator.

Now that we're happy this has worked, we can delete the old `variable` column:

```{r}
gapLong[,variable := NULL]
```

Finally, to split out the `value` column into the groups stored in `var`, we use
the `dcast` function. Since we'd like to keep the result as a data table, we'll
explicitly call the method from the `data.table` package:

```{r}
gapLong <- dcast.data.table(
  data=gapLong,  
  # unique identifier columns go to the left of the '~', separated by '+' signs.
  # The grouping column goes to the right of the '~'.
  formula=continent+country+year~var,
  # which column stores the values to be spread over the new columns
  value.var="value" 
)
```

And now we're back where we've started with the long data!

We've shown you the simplest cases for reshaping data. More complex cases may
require you to `melt` or `dcast` in several steps, and `rbind` the results.

Other packages you may find useful for this task are 
[tidyr](http://cran.r-project.org/web/packages/tidyr/index.html) and 
[splitstackshape](http://cran.r-project.org/web/packages/splitstackshape/index.html)
