---
layout: page
title: Intermediate R for reproducible scientific analysis
subtitle: For loops
minutes: 20
---

```{r, include=FALSE}
source("tools/chunk-options.R")
library("data.table")
gap <- fread("data/gapminder-FiveYearData.csv")
```
> ## Learning objectives {.objectives}
>
> * Write and understand `for` loops.
>

### Repeating operations

Often when we're trying to solve a problem or run some analysis we find 
ourselves doing the same thing over, and over, and over again on different 
groupings of data, or on different files, or with slight parameter variations.

The great thing about R, and programming in general, is it allows us to be 
**lazy**. Why do a repetitive task if you can make the computer do it for you?

For example, lets say I wanted to calculated the total population for each 
continent in the gapminder dataset in 2007. We could do this in several ways,
but the most basic approach is manually:

```{r}
gap[year == 2007 & continent == "Asia", sum(pop)]
gap[year == 2007 & continent == "Africa", sum(pop)]
gap[year == 2007 & continent == "Americas", sum(pop)]
gap[year == 2007 & continent == "Europe", sum(pop)]
gap[year == 2007 & continent == "Oceania", sum(pop)]
```

This is tedious to type out. We can do it, but imagine if we wanted
to run some calculation for each country!

The clever way to do this would be to use our recently acquired data.table 
skills:

```{r}
gap[year == 2007, sum(pop), by=continent]
```

But sometimes the solution to a problem isn't obvious, or doesn't fit into a
format we're used to, so it's helpful to have multiple tools in our 
problem-solving toolbox to fall back on. 

With a for loop we can instead *iterate* over each continent, and tell R to run
the same command:

```{r}
for (cc in gap[,unique(continent)]) {
  popsum <- gap[year == yy & continent == cc, sum(pop)]
  print(paste(cc, ":", popsum))
}
```

This construct tells R to go through each thing on the right of the `in` 
operator and store it in the variable `cc`. Inside the *body* of the `for` loop,
i.e. any lines of code that fall between the curly braces (`{` and `}`), we can 
then access the value of `cc` to do whatever we like. So first, `cc` will
hold the value "Asia", then it will run the line of code, and return back to the
top of the loop. Next `cc` will hold the value "Europe", and do the same thing,
and so on. 

What if we want to look at the change in total population for each continent
over the years? We can "nest" for loops to iterate through multiple separate
conditions:

```{r}
for (cc in gap[,unique(continent)]) {
  for (yy in gap[,unique(year)]) {
    popsum <- gap[year == yy & continent == cc, sum(pop)]
    print(paste(cc, yy, ":", popsum))
  }
}
```

#### For or Apply? The second circle of hell.

> We made our way into the second Circle, here live the gluttons.
> -- [The R inferno](http://www.burns-stat.com/pages/Tutor/R_inferno.pdf)

One of the biggest things that trips up novices and experienced R users alike, 
is building a results object (vector, list, matrix, data frame) as your for 
loop progresses. For example:

```{r}
results <- data.frame(continent=character(), year=numeric(), popsum=numeric())
for (cc in gap[,unique(continent)]) {
  for (yy in gap[,unique(year)]) {
    popsum <- gap[year == yy & continent == cc, sum(pop)]
    this_result <- data.frame(continent=cc, year=yy, popsum=popsum)
    results <- rbind(results, this_result)
  }
}
results
```

"Growing" a results object like this is bad practice. At each iteration, R needs
to talk to the computer's operating system to ask for the right amount of memory
for your new results object. Like all diplomatic negotiations, this can take a 
while (at least in computer time!). As a result, you might find that your for 
loops seem to take forever when you start working with bigger datasets or more
complex calculations.

It's much better to tell R how big your results object will be up front, that 
way R only needs to ask the computer for the right amount of memory once:

```{r}
# First lets calculate the number of rows we need:
nresults <- gap[,length(unique(continent))] * gap[,length(unique(year))] 
results <- data.frame(
  continent=character(length=nresults), 
  year=numeric(length=nresults), 
  popsum=numeric(length=nresults)
)
# Instead of iterating over values, we need to keep track of indices so we know
# which row to insert or new results into at each iteration. 
# `seq_along` will create a sequence of numbers based on the length of the 
# vector. So instead of c("Asia", "Americas", "Europe", "Africa", "Oceania"),
# ii will store c(1,2,3,4,5)
continents <- gap[,unique(continent)]
years <- gap[,unique(year)]
# We also need to keep track of which row to insert into. We could do fancy 
# math based on our indices, but this is hard to get right and can lead to hard
# to detect errors. Its much easier to just keep track of this manually. 
this_row <- 1
for (ii in seq_along(continents)) {
  for (jj in seq_along(years)) {
    # Now we need to look-up the appopriate values based on our indices
    cc <- continents[ii]
    yy <- years[jj]
    popsum <- gap[year == yy & continent == cc, sum(pop)]
    results[this_row,] <- list(cc, yy, popsum)
    # Increment the row counter
    this_row <- this_row + 1
  }
}
results
```

As you can see, this involves a lot more work. Most R users will even go so far
to tell you that for loops are bad, and that you should use something called
`apply` instead! We'll cover this in the next lesson, and later we'll show you
another method, `foreach` which also handles object creation for you.

For loops are most useful when you're performing a series of calculations where
each iteration depends on the results of the last (for example a random walk).


> #### Tip: While loops {.callout}
>
>
> Sometimes you will find yourself needing to repeat an operation until a certain
> condition is met. You can do this with a `while` loop.
> 
> ```{r, eval=FALSE}
> while(this condition is true){
>   do a thing
> }
> ```
> 
> As an example, here's a while loop 
> that generates random numbers from a uniform distribution (the `runif` function)
> between 0 and 1 until it gets one that's less than 0.1.
> 
> ~~~ {.r}
> z <- 1
> while(z > 0.1){
>   z <- runif(1)
>   print(z)
> }
> ~~~
> 
> `while` loops will not always be appropriate. You have to be particularly careful
> that you don't end up in an infinite loop because your condition is never met.
>

> #### Challenge 1 {.challenge}
>
> Write a script that loops through the `gapminder` data by continent and prints 
> out the mean life expectancy in 1952.
>

> #### Challenge 3 {.challenge}
>
> Modify the script so that it loops through the years as well as the continents.
>

> #### Challenge 4 {.challenge}
>
> Write a for loop that performs a random walk for 100 steps, then plot the 
> result.
>
> Hint: You can use `sign(rnorm(1))` in the body of the loop to randomly choose 
> a direction (forward or backward) at each iteration. 
>
> Hint: You will want to store the resulting position (starting at 0) after each
> iteration for plotting purposes.
>
> Hint: give the `plot` function the indices 0:100 as the x axis, and the 
> stored positions as the y axis. specify the 'type' argument as "l" to draw
> a the path.
>

