---
layout: page
title: IntermediateR for reproducible scientific analysis
subtitle: For loops
minutes: 20
---

```{r, include=FALSE}
source("tools/chunk-options.R")
# Silently load in the data so the rest of the lesson works
gapminder <- read.csv("data/gapminder-FiveYearData.csv", header=TRUE)
# Silently set seed for random number generation, so we don't have to explain it
set.seed(10)
```
> ## Learning objectives {.objectives}
>
> * Write and understand `for` loops.
>

### Repeating operations

 

If you want to iterate over
a set of values, when the order of iteration is important, and perform the
same operation on each, a `for` loop will do the job.
We saw `for` loops in the shell lessons earlier. This is the most
flexible of looping operations, but therefore also the hardest to use
correctly. Avoid using `for` loops unless the order of iteration is important:
i.e. the calculation at each iteration depends on the results of previous iterations.

The basic structure of a `for` loop is:

```{r, eval=FALSE}
for(iterator in set of values){
  do a thing
}
```

For example:

```{r}
for(i in 1:10){
  print(i)
}
```

The `1:10` bit creates a vector on the fly; you can iterate
over any other vector as well.

We can use a `for` loop nested within another `for` loop to iterate over two things at
once.

```{r}
for (i in 1:5){
  for(j in c('a', 'b', 'c', 'd', 'e')){
    print(paste(i,j))
  }
}
```

Rather than printing the results, we could write the loop output to a new object.

```{r}
output_vector <- c()
for (i in 1:5){
  for(j in c('a', 'b', 'c', 'd', 'e')){
    temp_output <- paste(i, j)
    output_vector <- c(output_vector, temp_output)
  }
}
output_vector
```

This approach can be useful, but 'growing your results' (building
the result object incrementally) is computationally inefficient, so avoid
it when you are iterating through a lot of values.

> #### Tip: don't grow your results {.callout}
>
> One of the biggest things that trips up novices and
> experienced R users alike, is building a results object
> (vector, list, matrix, data frame) as your for loop progresses.
> Computers are very bad at handling this, so your calculations
> can very quickly slow to a crawl. It's much better to define
> an empty results object before hand of the appropriate dimensions.
> So if you know the end result will be stored in a matrix like above,
> create an empty matrix with 5 row and 5 columns, then at each iteration
> store the results in the appropriate location.
>

A better way is to define your (empty) output object before filling in the values.
For this example, it looks more involved, but is still more efficient.

```{r}
output_matrix <- matrix(nrow=5, ncol=5)
j_vector <- c('a', 'b', 'c', 'd', 'e')
for (i in 1:5){
  for(j in 1:5){
    temp_j_value <- j_vector[j]
    temp_output <- paste(i, temp_j_value)
    output_matrix[i, j] <- temp_output
  }
}
output_vector2 <- as.vector(output_matrix)
output_vector2
```

> #### Tip: While loops {.callout}
>
>
> Sometimes you will find yourself needing to repeat an operation until a certain
> condition is met. You can do this with a `while` loop.
> 
> ```{r, eval=FALSE}
> while(this condition is true){
>   do a thing
> }
> ```
> 
> As an example, here's a while loop 
> that generates random numbers from a uniform distribution (the `runif` function)
> between 0 and 1 until it gets one that's less than 0.1.
> 
> ~~~ {.r}
> z <- 1
> while(z > 0.1){
>   z <- runif(1)
>   print(z)
> }
> ~~~
> 
> `while` loops will not always be appropriate. You have to be particularly careful
> that you don't end up in an infinite loop because your condition is never met.
>

> #### Challenge 3 {.challenge}
>
> Write a script that loops through the `gapminder` data by continent and prints out
> whether the mean life expectancy is smaller or larger than 50
> years.
>

> #### Challenge 4 {.challenge}
>
> Modify the script from Challenge 4 to also loop over each
> country. This time print out whether the life expectancy is
> smaller than 50, between 50 and 70, or greater than 70.
>

> #### Challenge 5 - Advanced {.challenge}
>
> Write a script that loops over each country in the `gapminder` dataset,
> tests whether the country starts with a 'B', and graphs life expectancy
> against time as a line graph if the mean life expectancy is under 50 years.
>
